\pkg{hesim}: Health Economic Simulation Modeling and Decision Analysis

### 2020-12-23 manuscript.Rnw
Minor updates to framework section

%% -- Framework ----------------------------------------------------------------
\section{Framework} \label{sec:framework}
Economic models consist of a disease model, a utility model, and a set of cost models for each cost category. Model development proceeds in 4 steps. The first step is to setup the model by defining the model structure, target population, and treatment strategies of interest, and storing the relevant information in a \code{hesim_data} object.

Development and analysis of the model is performed in the next three steps as shown in Figure~\ref{fig:econ-eval-process-hesim}. First, the disease progression, costs, and utility submodels are parameterized using ``estimation'' datasets. Next, the submodels are combined to construct an economic model. The economic model is then used to simulate, disease progression, QALYs, and costs as a function of ``input data''. The input data always contains variables describing the target population and treatment strategies of interest, but can also contain variables related to time to facilitate use of time-varying covariates or parameters. Finally, the simulated outcomes are used to perform decision analysis within a cost-effectiveness framework. While other approaches such as multi-criteria decision analysis (MCDA) could, in principle, be used as well, only CEA is currently supported. All models are simulated using PSA so that decision uncertainty can be represented.

@@

\subsection{Parameterization} \label{sec:parameterization}
Each submodel contains fields for the model parameters and the input data. Parameters can either be created from a statistical model fit using \proglang{R} or input directly by the user. There are two types of parameter objects, standard parameter objects prefixed by \code{``params''} and transformed parameter objects prefixed by \code{``tparams''}. The former contain the underlying parameters of a statistical model while the latter contain parameters more immediate to prediction that have been transformed as function of the input data. The regression coefficients of a logistic regression are example of a parameter objects while the predicted probabilities are examples of a transformed parameter object.

Transformed parameter objects are often useful when parameterizing a model using external sources. Two common examples are predicted probabilities for transition matrices and predicted means for assigning values to health states. The advantage of a transformed parameter object is that an explicit statistical model does not need to be specified which affords more flexibility to the user.

@@

The parameters of a survival model are stored in a \code{params_surv} object and a \code{params_surv_list} can be used to store the parameters of multiple survival models. The latter is useful for storing the parameters of a multi-state model or the independent survival models required for a PSM. The parameters of a multinomial logistic regression are stored in a \code{params_mlogit} and can be created by fitting a model with \code{nnet::multinom()}.

\code{tparams_transprobs} objects contain transition probability matrices that have been predicted for each PSA sample, treatment strategy, and patient from the target population. Transition probabilities may also vary by time interval to allow for time inhomogeneous Markov models. Transition probabilities can be predicted from a fitted multi-state model using the \pkg{msm} package or constructed ``by hand'' in a custom manner.

\subsubsection{Costs and utility} 
State values (i.e., costs and utilities) do not depend on the choice of disease model. They can currently either be modeled using a linear model or using predicted means.

\begin{table} [h]
\caption{Parameterization of state value models}\label{tbl:parameterize-disease-model}
\footnotesize
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}l l l l}
\hline
Statistical model & Parameter object & Model object\\
\hline
Predicted means & \code{tparams_mean} & \code{stateval_tbl::lm}\\
Linear model & \code{params_lm} & \code{stats::lm}\\
\hline
\end{tabular*}
\end{table}

The most straightforward way to construct state values is with a \code{stateval_tbl}, which is a special object used to assign values (i.e. predicted means) to health states that can vary across PSA samples, treatment strategies, patients, and/or time intervals. State values can be specified either as moments (i.e., mean and standard error) or parameters (e.g., shape and scale of gamma distribution) of a probability distribution, or by pre-simulating values from a suitable probability distribution (e.g., from a Bayesian model).

@@

The disease and state value models are most easily instantiated using \code{S3} generic functions prefixed by ``\code{create}'' from parameter, transformed parameter, or statistical model objects. An \code{IndivCtstmTrans} can, for instance, be created from a \code{flexsurvreg_list} or \code{params_surv_list} object with \code{create_IndivCtstmTrans()}. Similarly, a \code{StateVals} object can, for example, be created from a \code{stateval_tbl} object with \code{create_StateVals()}. The complete economic model is instantiated by combining the submodels using the \pkg{R6} constructor method \code{$new()}.

Each economic model contains methods for simulating disease progression, QALYs, and costs (Table~\ref{tbl:simulate-outcomes}). The cost and utility models always simulate costs and QALYs from the simulated  progression of disease with the methods \code{$sim_qalys()} and \code{$sim_costs()}, respectively. Methods for simulating disease progression differ slightly since the simulation approaches differ as described in Section~\ref{sec:model-taxonomy}. In an N-state PSM $n-1$ survival curves are generated and in a iCTSTM a trajectory through state transitions is simulated for multiple patients via individual simulation. All models can simulate state probabilities over time.

>>>>

%% -- Framework ----------------------------------------------------------------
\section{Framework} \label{sec:framework}
Economic models consist of a disease model, a utility model, and a set of cost models for each cost category. Model development proceeds in 4 steps. The first step is to setup the model by defining the model structure, target population, and treatment strategies of interest. The relevant information is then stored in a \code{hesim_data} object.

Development and analysis of the model is performed in the next three steps as shown in Figure~\ref{fig:econ-eval-process-hesim}. First, the disease progression, costs, and utility submodels are parameterized using ``estimation'' datasets. Next, the submodels are combined to construct an economic model. The economic model is then used to simulate, disease progression, QALYs, and costs as a function of ``input data''. The input data always contains variables describing the target population and treatment strategies of interest, but can also contain variables related to time to facilitate use of time-varying covariates or parameters. Finally, the simulated QALYs and costs are used to perform decision analysis within a cost-effectiveness framework. While other approaches such as multi-criteria decision analysis (MCDA) could, in principle, be used as well, only CEA is currently supported. All models are simulated using PSA so that decision uncertainty can be represented.

@@

\subsection{Parameterization} \label{sec:parameterization}
Each submodel contains fields for the model parameters and the input data. Parameters can either be estimated from a statistical model fit using \proglang{R} or input directly by the user. There are two types of parameter objects, standard parameter objects prefixed by \code{``params''} and ``transformed'' parameter objects prefixed by \code{``tparams''}. The former contain the underlying parameters of a statistical model while the latter contain parameters more immediate to prediction that have been transformed as function of the input data. The regression coefficients of a logistic regression are an example of a parameter objects while the predicted probabilities are examples of a transformed parameter object.

Transformed parameter objects are often useful when parameterizing a model using external sources. Two common examples are the predicted probabilities comprising transition probability matrices and predicted means for assigning values to health states. The advantage of a transformed parameter object is that an explicit statistical model does not need to be specified which gives the user more flexibility.

@@

The parameters of a survival model are stored in a \code{params_surv} object and a \code{params_surv_list} can be used to store the parameters of multiple survival models. The latter is useful for storing the parameters of a multi-state model or the independent survival models required for a PSM. The parameters of a multinomial logistic regression are stored in a \code{params_mlogit} and can be created by fitting a model for each row in a transition probability matrix with \code{nnet::multinom()}.

\code{tparams_transprobs} objects store transition probability matrices that have been predicted for each PSA sample, treatment strategy, patient from the target population, and optionally time interval (to allow for time inhomogeneous Markov models). Transition probabilities can be predicted from a fitted multi-state model using the \pkg{msm} package or constructed ``by hand'' in a custom manner.

\subsubsection{Costs and utility} 
State values (i.e., costs and utilities) do not depend on the choice of disease model. They can currently either be modeled using a linear model or using predicted means (Table~\ref{tbl:parameterize-stateval-model}).

\begin{table} [h]
\caption{Parameterization of state value models}\label{tbl:parameterize-stateval-model}
\footnotesize
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}l l l l}
\hline
Statistical model & Parameter object & Model object\\
\hline
Predicted means & \code{tparams_mean} & \code{stateval_tbl::lm}\\
Linear model & \code{params_lm} & \code{stats::lm}\\
\hline
\end{tabular*}
\end{table}

The most straightforward way to construct state values is with a \code{stateval_tbl}, which is a special object used to assign values (i.e. predicted means) to health states that can vary across PSA samples, treatment strategies, patients, and/or time intervals. State values can be specified either as moments (e.g., mean and standard error) or parameters (e.g., shape and scale of gamma distribution) of a probability distribution, or by pre-simulating values from a suitable probability distribution (e.g., from a Bayesian model).

@@

The disease and state value models can either be instantiated from parameter or model objects using \code{S3} generic functions prefixed by ``\code{create}'' or using the \pkg{R6} constructor method \code{$new()}. \code{create_IndivCtstmTrans()} is an example of the former that can be used to create an \code{IndivCtstmTrans} object from a \code{flexsurvreg}, \code{flexsurvreg_list}, \code{params_surv}, or \code{params_surv_list} object. Similarly, a \code{StateVals} object can, for example, be created from a \code{stateval_tbl} object with \code{create_StateVals()}. The complete economic model is instantiated by combining the submodels using \code{$new()}.

Each economic model contains methods for simulating disease progression, QALYs, and costs (Table~\ref{tbl:simulate-outcomes}). The utility and cost models always simulate QALYs and costs from the simulated progression of disease with the methods \code{$sim_qalys()} and \code{$sim_costs()}, respectively. Methods for simulating disease progression differ slightly since the simulation approaches differ as described in Section~\ref{sec:model-taxonomy}. In an N-state PSM $n-1$ survival curves are generated and in a iCTSTM a trajectory through the STM is simulated for multiple patients via individual simulation. All models can simulate state probabilities over time.
